---
title: "fertilization_millet_yield"
author: "Douglas Lawton"
format: html
editor: source
---


# Question 1: How does fertilization influence millet yield

This data is a time series with three sample points (mission 1-3) and 200-ish farmers all gps referenced. Temperature and humidity are recorded as well. All these are variables that we want to control and are not necessarily concerned about.

## Library import and reading in data

```{r reading in packages, warning=FALSE,message=FALSE}


# Define a function to check and install packages if not already installed
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  }
}



# List of packages to install and load
packages <- c("tidyverse", "here", "janitor",  
              "ggpubr", "MetBrewer", "patchwork",'sf',
              'rnaturalearth','rnaturalearthdata','mgcv','gratia','emmeans')

# Call the function with your package list
install_and_load(packages)


options(width=200)
here::i_am('README.md')

```

```{r}

dat <- read_csv('data/analysis_ready_data.csv')

```

This should be self explanatory, but lets see what the data looks like.

```{r exploratory plots}

raw_yield_farmer_plot <- dat |>
  ggplot(aes(x=treatment,y=yield,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    theme_pubr() +
    ylab("Yield") +
    xlab("") +
    facet_wrap(~village)

raw_yield_plot <- dat |>
  ggplot(aes(x=treatment,y=yield)) +
    geom_jitter(width=0.2,height=0) +
    theme_pubr() +
    ylab("Yield") +
    xlab("") 

raw_yield_boxplot_plot <- dat |>
  ggplot(aes(x=treatment,y=yield)) +
    geom_boxplot(outlier.size = -1) +
    geom_jitter(width=0.2,height=0,pch=21,color='black') +
    theme_pubr() +
    ylab("Yield") +
    xlab("") 


ggsave(raw_yield_farmer_plot,
       file='output/raw_yield_plots/raw_yield_farmer_plot.png',
       width=10,height=5,dpi=300)

ggsave(raw_yield_plot,
       file='output/raw_yield_plots/raw_yield_plot.png',
       width=5,height=5,dpi=300)

ggsave(raw_yield_boxplot_plot,
       file='output/raw_yield_plots/raw_yield_boxplot_plot.png',
       width=5,height=5,dpi=300)


raw_yield_farmer_plot
raw_yield_plot
raw_yield_boxplot_plot
```

So pretty obvious, fertilizer influences yield. It doesn't appear to be extreme differences between village (with the exception of some farmers). Some farmers had considerably more yield increase than others which could be contributed to many variables (rainfall, soils, land management etc.)


Lets build a statistical model with village and farmer as random effects and look directly at the impact of fertilization on yield. Again, there will be no surprise here. I'll do some model selection just to see variable importance (just for kicks)

```{r yield modeling}

yield_mod_dat <- dat |>
  select(yield,treatment,village,farmer_num,village) |>
  mutate(village = factor(village),
    treatment = factor(treatment)) |>
  distinct() |> 
  drop_na(yield)
  
  
mod0 <- gam(yield ~ 1,
            data=yield_mod_dat,family = tw()) # null model

mod1 <- gam(yield ~ treatment,
            data=yield_mod_dat,family = tw()) # basic ANOVA

mod2 <- gam(yield ~ treatment + 
              s(village,bs='re'),
            data=yield_mod_dat,family = tw())

mod3 <- gam(yield ~ treatment +
              s(farmer_num,bs='re'),
            data=yield_mod_dat,family = tw())

mod4 <- gam(yield ~ treatment + 
              s(village,bs='re') + 
              s(farmer_num,bs='re'),
            data=yield_mod_dat,family = tw())

```

Alright, lets check the models for fit and select via AIC:

```{r model checking}

gratia::draw(mod3,parametric = TRUE)

AIC(mod0,mod1,mod2,mod3,mod4) |>
  rownames_to_column('model') |>
  as_tibble() |>
  mutate(deltaAIC = AIC - min(AIC)) |>
  arrange(deltaAIC)
```

okay interesting! Village did not a lot of variance in yield, but farmer ID did. This should make sense for our exploratory plots as there is considerable differences between the paired plot shifts between farmer.

With this said, the effect on treatment overall was small if you compare the parametric estimate between `mod1` (0.81) and `mod3` (0.88):

```{r yield mod1 and mod3 summary}

summary(mod1)
summary(mod3)

```

Lets visually plot the final estimated marginal means:\

```{r emm plotting}

mod1_emm <- emmeans(mod1,'treatment',type='response') |>
  as_tibble() |>
  mutate(model = 'ANOVA')

mod3_emm <- emmeans(mod3,'treatment',type='response') |>
  as_tibble() |>
  mutate(model = 'GLMM')

emms <- rbind(mod1_emm,mod3_emm)

yield_anova_glmms_estimated_marginal_means_comparisons <- emms |>
  ggplot(aes(x=treatment,y=response,color=model)) +
    geom_point(size=5,position = position_dodge(width=0.4)) +
    geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                  width=0,position = position_dodge(width=0.4)) +
    theme_pubr() +
    ylab("Yield") +
    xlab("") +
    scale_color_met_d(name='Degas')

yield_anova_glmms_estimated_marginal_means_comparisons

ggsave(yield_anova_glmms_estimated_marginal_means_comparisons,
       file='output/modeled_yield_plots/yield_anova_glmms_estimated_marginal_means_comparisons.png',
       width=5,height=5,dpi=300)
```

So the final plot looks like this:

```{r final yield model plot}

final_yield_model_est_marginal_means <- emms |>
  filter(model == 'GLMM') |>
  ggplot(aes(x=treatment,y=response)) +
    geom_point(size=5,position = position_dodge(width=0.4)) +
    geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                  width=0,position = position_dodge(width=0.4)) +
    theme_pubr() +
    ylab("Yield") +
    xlab("") +
    scale_color_met_d(name='Degas')


farmer_yield_variance <- smooth_estimates(mod3) |>
  ggplot(aes(x=reorder(farmer_num,-.estimate),y=.estimate)) +
    geom_point() +
    geom_errorbar(aes(ymin=.estimate-.se,ymax=.estimate+.se),width=0) +
    coord_flip() +
    xlab("Farmer Number") +
    ylab("yield estimate") +
    theme_pubr()

ggsave(final_yield_model_est_marginal_means,
       file='output/modeled_yield_plots/final_yield_model_est_marginal_means.png',
       width=5,height=5,dpi=300)

ggsave(farmer_yield_variance,
       file='output/modeled_yield_plots/farmer_yield_variance.png',
       width=5,height=12,dpi=300)

final_yield_model_est_marginal_means
farmer_yield_variance

```


## Saving to disk for figure and table construction:


```{r}

yield_data <- dat |>
  select(farmer_num,village,treatment,yield) |>
  left_join(emms,by='treatment')


write.csv(yield_data,file=here('data/processed/yield_raw_and_model_data.csv'))


```