---
title: "locust abundance and attack"
author: "Douglas Lawton"
format: html
editor: source
---


## Question 2: Does fertilization influence *O. senegalensis* abundance and attack

This data is a time series with three sample points (mission 1-3) and 200-ish farmers all gps referenced. Temperature and humidity are recorded as well. All these are variables that we want to control and are not necessarily concerned about.

## Library import and reading in data

```{r reading in packages, warning=FALSE,message=FALSE}


# Define a function to check and install packages if not already installed
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  }
}



# List of packages to install and load
packages <- c("tidyverse", "here", "janitor",  
              "ggpubr", "MetBrewer", "patchwork",'sf',
              'rnaturalearth','rnaturalearthdata','mgcv','gratia','emmeans')

# Call the function with your package list
install_and_load(packages)


options(width=200)
here::i_am('README.md')

```


## Raw data visualization

Lets also get a time of day measure as this can affect locust abundances greatly.


```{r}

dat <- read_csv('data/analysis_ready_data.csv') |>
  mutate(hour = lubridate::hour(datetime)) |>
  filter(survey != 's1')

```

```{r OSE exploratory model}

# Basic plots
# List of plots with their respective file paths
plot_list <- list(
  raw_ose_count_treatment = dat |>
    select(village,farmer_num,treatment,ose_num_adj) |>
    distinct() |>
    ggplot(aes(x = treatment, y = ose_num_adj)) +
    geom_jitter(width = 0.2) +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr(),
  
  raw_ose_count_humidity_percent = dat |>
    select(village,farmer_num,treatment,ose_num_adj,humidity_percent) |>
    distinct() |>
    ggplot(aes(x = humidity_percent, y = ose_num_adj)) +
    geom_jitter(width = 0.2) +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr(),
  
  raw_ose_count_ground_cover_percent = dat |>
    select(village,farmer_num,treatment,ose_num_adj,ground_cover_percent) |>
    distinct() |>
    ggplot(aes(x = ground_cover_percent, y = ose_num_adj)) +
    geom_jitter(width = 0.2) +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr(),
  
  raw_ose_count_hour = dat |>
    select(village,farmer_num,treatment,ose_num_adj,hour) |>
    distinct() |>
    ggplot(aes(x = hour, y = ose_num_adj)) +
    geom_jitter(width = 0.2) +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr(),
  
  raw_millet_attack_treatment = dat |>
    select(village,farmer_num,treatment,att_mil_adj) |>
    distinct() |>
    ggplot(aes(x = treatment, y = att_mil_adj)) +
    geom_jitter(width = 0.2) +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr(),
  
  raw_millet_attack_humidity_percent = dat |>
    select(village,farmer_num,treatment,humidity_percent,att_mil_adj) |>
    distinct() |>
    ggplot(aes(x = humidity_percent, y = att_mil_adj)) +
    geom_point()  +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr(),
  
  raw_millet_attack_ground_cover_percent = dat |>
    select(village,farmer_num,treatment,ground_cover_percent,att_mil_adj) |>
    distinct() |>
    ggplot(aes(x = ground_cover_percent, y = att_mil_adj)) +
    geom_point() +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr(),
  
  raw_millet_attack_hour = dat  |>
    select(village,farmer_num,treatment,hour,att_mil_adj) |>
    distinct() |>
    ggplot(aes(x = hour, y = att_mil_adj)) +
    geom_point() +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr()
)

# File paths
file_paths <- list(
  raw_ose_count_treatment = 'output/ose_treatment_plots/raw_ose_count_treatment.png',
  raw_ose_count_humidity_percent = 'output/ose_treatment_plots/raw_ose_count_humidity_percent.png',
  raw_ose_count_ground_cover_percent = 'output/ose_treatment_plots/raw_ose_count_ground_cover_percent.png',
  raw_ose_count_hour = 'output/ose_treatment_plots/raw_ose_count_hour.png',
  raw_millet_attack_treatment = 'output/ose_treatment_plots/raw_millet_attack_treatment.png',
  raw_millet_attack_humidity_percent = 'output/ose_treatment_plots/raw_millet_attack_humidity_percent.png',
  raw_millet_attack_ground_cover_percent = 'output/ose_treatment_plots/raw_millet_attack_ground_cover_percent.png',
  raw_millet_attack_hour = 'output/ose_treatment_plots/raw_millet_attack_hour.png'
)

# Save plots using a loop
for (plot_name in names(plot_list)) {
  ggsave(
    plot_list[[plot_name]],
    file = file_paths[[plot_name]],
    width = 5, height = 5, dpi = 300
  )
}

```

So the pattern isnt as strong as yield, but I think there is still something there. Lets break it down by village, farmer, and mission.

```{r breaking plots down by village farmer and mission for locust abundance}

# Broken out into village and farmer

ose_number_village_treatment <- dat  |>
  select(village,farmer_num,treatment,ose_num_adj) |>
  distinct() |>
  ggplot(aes(x=treatment,y=ose_num_adj,color=village)) +
    geom_jitter(width=0.2) +
    scale_color_met_d(name='Demuth') +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr()


ose_number_village_survey <- dat |>
  filter(survey != 's1') |>
  select(village,survey,farmer_num,treatment,ose_num_adj) |>
  distinct() |>
  drop_na() |>
  ggplot(aes(x=treatment,y=ose_num_adj,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr() +
    facet_wrap(~survey)


# What about across survey?

ose_number_treatment_village_survey <- dat |>
  select(village,survey,farmer_num,treatment,ose_num_adj) |>
  distinct() |>
  drop_na() |>
  ggplot(aes(x=survey,y=ose_num_adj,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    theme_pubr() +
    facet_wrap(~treatment)



ggsave(ose_number_village_treatment,
       file='output/ose_treatment_plots/ose_number_village_treatment.png',
       width=5,height=5,dpi=300)

ggsave(ose_number_village_survey,
       file='output/ose_treatment_plots/ose_number_village_survey.png',
       width=10,height=5,dpi=300)

ggsave(ose_number_treatment_village_survey,
       file='output/ose_treatment_plots/ose_number_treatment_village_survey.png',
       width=10,height=5,dpi=300)


ose_number_village_treatment
ose_number_village_survey
ose_number_treatment_village_mission

```

## Locust abundance and fertilizer treatment

lets make models regressing locust abundance on treatment.

```{r modeling locust abundance}

dat <- dat |>
  mutate(across(where(is.character),~as.factor(.)))


s1_exclude <- dat |>
  filter(survey != 's1')


ose_num_mod0 <- gam(ose_num_adj ~ 1,
            data=s1_exclude,family = tw()) # null model

ose_num_mod1 <- gam(ose_num_adj ~ treatment,
            data=s1_exclude,family = tw()) # basic ANOVA

ose_num_mod2 <- gam(ose_num_adj ~ treatment + 
              s(survey,bs='re') +
              s(village,bs='re'),
            data=s1_exclude,family = tw())

ose_num_mod3 <- gam(ose_num_adj ~ treatment +
              s(survey,bs='re') +
              s(farmer_num,bs='re'),
            data=s1_exclude,family = tw())

ose_num_mod4 <- gam(ose_num_adj ~ treatment + 
              survey +
              s(village,bs='re') + 
              s(farmer_num,bs='re'),
            data=s1_exclude,family = tw())

AIC(ose_num_mod0,ose_num_mod1,ose_num_mod2,ose_num_mod3,ose_num_mod4)

summary(ose_num_mod4)
```

### creating table output of model results

```{r}

# Extract parametric and smooth terms
param_table <- ose_num_mod4 |> broom::tidy(parametric = TRUE)
smooth_table <- ose_num_mod4 |> broom::tidy(parametric = FALSE)

# Standardize column names for merging
smooth_table <- smooth_table |> 
  rename(std.error = edf, estimate = ref.df) |>  # Renaming for consistency
  mutate(std.error = NA)  # Adding missing column to match parametric table

# Combine both tables
combined_table <- bind_rows(param_table, smooth_table) |> 
  mutate(type = c(rep("Parametric", nrow(param_table)), rep("Non-parametric", nrow(smooth_table))),
    term = gsub('treatment|survey','',term),
    across(where(is.numeric), ~ round(.,3)))

write_csv(combined_table,file=here('output/model_summary_files/locust_count_treatment_model_results.csv'))

```


## locust abundance and millet attack
Now lets do the same thing for millet attack:

```{r breaking plots down by village farmer and mission for locust abundance}

# Broken out into village and farmer

dat |>
  ggplot(aes(x=treatment,y=att_mil_adj,color=village)) +
    geom_jitter(width=0.2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr()


dat |>
  filter(survey != 's4') |>
  ggplot(aes(x=treatment,y=att_mil_adj,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~survey,ncol=2)

# What about across survey?

dat |>
  filter(survey != 's4') |>
  ggplot(aes(x=survey,y=att_mil_adj,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~treatment)


# Okay what about yield

dat |>
  filter(survey != 's4') |>
  ggplot(aes(x=survey,y=att_mil_adj,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~treatment)

```

```{r millet attack}


att_mil_mod0 <- gam(att_mil_adj ~ 1,
            data=s1_exclude,family = tw()) # null model

att_mil_mod1 <- gam(att_mil_adj ~ treatment,
            data=s1_exclude,family = tw()) # basic ANOVA

att_mil_mod2 <- gam(att_mil_adj ~ treatment + 
              s(survey,bs='re') +
              s(village,bs='re'),
            data=s1_exclude,family = tw())

att_mil_mod3 <- gam(att_mil_adj ~ treatment +
              s(survey,bs='re') +
              s(farmer_num,bs='re'),
            data=s1_exclude,family = tw())

att_mil_mod4 <- gam(att_mil_adj ~ treatment + 
              s(survey,bs='re') +
              s(village,bs='re') + 
              s(farmer_num,bs='re'),
            data=s1_exclude,family = tw())

AIC(att_mil_mod0,att_mil_mod1,att_mil_mod2,att_mil_mod3,att_mil_mod4)
summary(att_mil_mod4)
appraise(att_mil_mod1)
```



### saving millet attack model summary

```{r}

# Extract parametric and smooth terms
param_table <- att_mil_mod4 |> broom::tidy(parametric = TRUE)
smooth_table <- att_mil_mod4 |> broom::tidy(parametric = FALSE)

# Standardize column names for merging
smooth_table <- smooth_table |> 
  rename(std.error = edf, estimate = ref.df) |>  # Renaming for consistency
  mutate(std.error = NA)  # Adding missing column to match parametric table

# Combine both tables
combined_table <- bind_rows(param_table, smooth_table) |> 
  mutate(type = c(rep("Parametric", nrow(param_table)), rep("Non-parametric", nrow(smooth_table))),
    term = gsub('treatment','',term),
    across(where(is.numeric), ~ round(.,3)))

write_csv(combined_table,file=here('output/model_summary_files/locust_attack_treatment_model_results.csv'))

```

## plotting model estimates


There are some issues with the model diagnostics, but not too bad. Lets summarize this all up:

```{r abundance and millet attack final plots}

ose_num_emms <- emmeans(ose_num_mod4,'treatment',type='response') |> as_tibble()

att_mil_emms <- emmeans(att_mil_mod4,'treatment',type='response') |> as_tibble()

ose_abundance_modeled <- ose_num_emms |>
  as_tibble() |>
  ggplot(aes(x=treatment,y=response)) +
    geom_point(size=5) +
    geom_errorbar(aes(ymin = lower.CL,ymax = upper.CL),width=0) +
    theme_pubr() +
    ylab("locust abundance") 

ose_millet_attack_modeled <- att_mil_emms |>
  as_tibble() |>
  ggplot(aes(x=treatment,y=response)) +
    geom_point(size=5) +
    geom_errorbar(aes(ymin = lower.CL,ymax = upper.CL),width=0)  +
    theme_pubr()  +
    ylab("locust damage") 

ose_abundance_modeled
ose_millet_attack_modeled

ggsave(ose_abundance_modeled,
       file='output/ose_treatment_plots/ose_abundance_modeled.png',
       width=5,height=5,dpi=300)

ggsave(ose_abundance_modeled,
       file='output/ose_treatment_plots/ose_abundance_modeled.png',
       width=5,height=5,dpi=300)

```

and the model summaries:

```{r model summaries}

summary(ose_num_mod4)

summary(att_mil_mod4)
```


## Saving to disk for figure and table construction:


```{r}

locust_dat <- dat |>
  select(farmer_num,village,treatment,ose_num_adj,att_mil_adj,survey) |>
  left_join(ose_num_emms,by='treatment') |>
  rename(abundance_response = response,abundance_se = SE,abundance_df = df, abundance_lower.CL = lower.CL, abundance_upper.CL = upper.CL) |>
  left_join(att_mil_emms,by='treatment') |>
  rename(att_mil_response = response,att_mil_se = SE,att_mil_df = df, att_mil_lower.CL = lower.CL, att_mil_upper.CL = upper.CL)


write.csv(locust_dat,file=here('data/processed/locust_abundance_and_att_mil_raw_and_model_data.csv'))


```
