
## Question 2: Does fertilization influence *O. senegalensis* abundance and attack

This data is a time series with three sample points (mission 1-3) and 200-ish farmers all gps referenced. Temperature and humidity are recorded as well. All these are variables that we want to control and are not necessarily concerned about.

## Library import and reading in data

```{r reading in packages, warning=FALSE,message=FALSE}


# Define a function to check and install packages if not already installed
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  }
}



# List of packages to install and load
packages <- c("tidyverse", "here", "janitor",  
              "ggpubr", "MetBrewer", "patchwork",'sf',
              'rnaturalearth','rnaturalearthdata','mgcv','gratia','emmeans')

# Call the function with your package list
install_and_load(packages)


options(width=200)
here::i_am('README.md')

```


## Raw data visualization

Lets also get a time of day measure as this can affect locust abundances greatly.


```{r}

dat <- dat |>
  mutate(hour = lubridate::hour(datetime))
names(dat)
```

```{r OSE exploratory model}

# Basic plots
# List of plots with their respective file paths
plot_list <- list(
  raw_ose_count_treatment = dat |>
    ggplot(aes(x = treatment, y = ose_num)) +
    geom_jitter(width = 0.2) +
    theme_pubr(),
  
  raw_ose_count_humidity_percent = dat |>
    ggplot(aes(x = humidity_percent, y = ose_num)) +
    geom_jitter(width = 0.2) +
    theme_pubr(),
  
  raw_ose_count_ground_cover_percent = dat |>
    ggplot(aes(x = ground_cover_percent, y = ose_num)) +
    geom_jitter(width = 0.2) +
    theme_pubr(),
  
  raw_ose_count_hour = dat |>
    ggplot(aes(x = hour, y = ose_num)) +
    geom_jitter(width = 0.2) +
    theme_pubr(),
  
  raw_millet_attack_treatment = dat |>
    ggplot(aes(x = treatment, y = att_mil)) +
    geom_jitter(width = 0.2) +
    theme_pubr(),
  
  raw_millet_attack_humidity_percent = dat |>
    ggplot(aes(x = humidity_percent, y = att_mil)) +
    geom_point() +
    theme_pubr(),
  
  raw_millet_attack_ground_cover_percent = dat |>
    ggplot(aes(x = ground_cover_percent, y = att_mil)) +
    geom_point() +
    theme_pubr(),
  
  raw_millet_attack_hour = dat |>
    ggplot(aes(x = hour, y = att_mil)) +
    geom_point() +
    theme_pubr()
)

# File paths
file_paths <- list(
  raw_ose_count_treatment = 'output/ose_treatment_plots/raw_ose_count_treatment.png',
  raw_ose_count_humidity_percent = 'output/ose_treatment_plots/raw_ose_count_humidity_percent.png',
  raw_ose_count_ground_cover_percent = 'output/ose_treatment_plots/raw_ose_count_ground_cover_percent.png',
  raw_ose_count_hour = 'output/ose_treatment_plots/raw_ose_count_hour.png',
  raw_millet_attack_treatment = 'output/ose_treatment_plots/raw_millet_attack_treatment.png',
  raw_millet_attack_humidity_percent = 'output/ose_treatment_plots/raw_millet_attack_humidity_percent.png',
  raw_millet_attack_ground_cover_percent = 'output/ose_treatment_plots/raw_millet_attack_ground_cover_percent.png',
  raw_millet_attack_hour = 'output/ose_treatment_plots/raw_millet_attack_hour.png'
)

# Save plots using a loop
for (plot_name in names(plot_list)) {
  ggsave(
    plot_list[[plot_name]],
    file = file_paths[[plot_name]],
    width = 5, height = 5, dpi = 300
  )
}

```

So the pattern isnt as strong as yield, but I think there is still something there. Lets break it down by village, farmer, and mission.

```{r breaking plots down by village farmer and mission for locust abundance}

# Broken out into village and farmer

ose_number_village_treatment <- dat |>
  ggplot(aes(x=treatment,y=ose_num,color=village)) +
    geom_jitter(width=0.2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr()


ose_number_village_mission <- dat |>
  ggplot(aes(x=treatment,y=ose_num,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~mission,ncol=2)

# What about across mission?

ose_number_treatment_village_mission <- dat |>
  ggplot(aes(x=mission,y=ose_num,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~treatment)



ggsave(ose_number_village_treatment,
       file='output/ose_treatment_plots/ose_number_village_treatment.png',
       width=5,height=5,dpi=300)

ggsave(ose_number_village_mission,
       file='output/ose_treatment_plots/ose_number_village_mission.png',
       width=10,height=10,dpi=300)

ggsave(ose_number_treatment_village_mission,
       file='output/ose_treatment_plots/ose_number_treatment_village_mission.png',
       width=10,height=5,dpi=300)


ose_number_village_treatment
ose_number_village_mission
ose_number_treatment_village_mission

```

lets make models regressing locust abundance on treatment.

```{r modeling locust abundance}

m1_exclude <- dat |>
  filter(mission != 'm1')


ose_num_mod0 <- gam(ose_num ~ 1,
            data=m1_exclude,family = tw()) # null model

ose_num_mod1 <- gam(ose_num ~ treatment,
            data=m1_exclude,family = tw()) # basic ANOVA

ose_num_mod2 <- gam(ose_num ~ treatment + 
              s(mission,bs='re') +
              s(village,bs='re'),
            data=m1_exclude,family = tw())

ose_num_mod3 <- gam(ose_num ~ treatment +
              s(mission,bs='re') +
              s(farmer_num,bs='re'),
            data=m1_exclude,family = tw())

ose_num_mod4 <- gam(ose_num ~ treatment + 
              mission +
              s(village,bs='re') + 
              s(farmer_num,bs='re'),
            data=m1_exclude,family = tw())

AIC(ose_num_mod0,ose_num_mod1,ose_num_mod2,ose_num_mod3,ose_num_mod4)

summary(ose_num_mod4)
```

Now lets do the same thing for millet attack:

```{r breaking plots down by village farmer and mission for locust abundance}

# Broken out into village and farmer

dat |>
  ggplot(aes(x=treatment,y=att_mil,color=village)) +
    geom_jitter(width=0.2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr()


dat |>
  ggplot(aes(x=treatment,y=att_mil,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~mission,ncol=2)

# What about across mission?

dat |>
  ggplot(aes(x=mission,y=att_mil,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~treatment)


# Okay what about yield

dat |>
  ggplot(aes(x=mission,y=att_mil,color=village,group=farmer_num)) +
    geom_line(alpha=0.5) +
    geom_point(size=2) +
    scale_color_met_d(name='Demuth') +
    theme_pubr() +
    facet_wrap(~treatment)

```

```{r millet attack}


att_mil_mod0 <- gam(att_mil ~ 1,
            data=dat,family = tw()) # null model

att_mil_mod1 <- gam(att_mil ~ treatment,
            data=dat,family = tw()) # basic ANOVA

att_mil_mod2 <- gam(att_mil ~ treatment + 
              s(mission,bs='re') +
              s(village,bs='re'),
            data=dat,family = tw())

att_mil_mod3 <- gam(att_mil ~ treatment +
              s(mission,bs='re') +
              s(farmer_num,bs='re'),
            data=dat,family = tw())

att_mil_mod4 <- gam(att_mil ~ treatment + 
              s(mission,bs='re') +
              s(village,bs='re') + 
              s(farmer_num,bs='re'),
            data=dat,family = tw())

AIC(att_mil_mod0,att_mil_mod1,att_mil_mod2,att_mil_mod3,att_mil_mod4)
summary(att_mil_mod4)
appraise(att_mil_mod1)
```

There are some issues with the model diagnostics, but not too bad. Lets summarize this all up:

```{r abundance and millet attack final plots}

ose_abundance_modeled <- emmeans(ose_num_mod4,'treatment',type='response') |>
  as_tibble() |>
  ggplot(aes(x=treatment,y=response)) +
    geom_point(size=5) +
    geom_errorbar(aes(ymin = lower.CL,ymax = upper.CL),width=0) +
    theme_pubr() +
    ylab("locust abundance") 

ose_millet_attack_modeled <- emmeans(att_mil_mod4,'treatment',type='response') |>
  as_tibble() |>
  ggplot(aes(x=treatment,y=response)) +
    geom_point(size=5) +
    geom_errorbar(aes(ymin = lower.CL,ymax = upper.CL),width=0)  +
    theme_pubr()  +
    ylab("locust damage") 

ose_abundance_modeled
ose_millet_attack_modeled

ggsave(ose_abundance_modeled,
       file='output/ose_treatment_plots/ose_abundance_modeled.png',
       width=5,height=5,dpi=300)

ggsave(ose_abundance_modeled,
       file='output/ose_treatment_plots/ose_abundance_modeled.png',
       width=5,height=5,dpi=300)

```

and the model summaries:

```{r model summaries}

summary(ose_num_mod4)

summary(att_mil_mod4)
```





```{r}
dat


mod <- gam(ose_num ~ treatment + 
              s(hour) +
              s(temp_c) +
              s(humidity_percent) +
              s(mission,bs='re') +
              s(village,bs='re'),
            data=dat,family = tw())

mod2 <- gam(ose_num ~ treatment + 
              s(mission,bs='re') +
              s(village,bs='re'),
            data=dat,family = tw())

mod <- gam(att_mil ~ treatment + 
              s(hour) +
              s(temp_c) +
              s(humidity_percent) +
              s(mission,bs='re') +
              s(village,bs='re'),
            data=dat,family = tw())

mod2 <- gam(att_mil ~ treatment + 
              s(mission,bs='re') +
              s(village,bs='re'),
            data=dat,family = tw())
      

summary(mod1)
gratia::appraise(mod)
```