---
title: "locust millet attack"
author: "Douglas Lawton"
format: html
editor: source
---

## Supplementary Question: Does *O. senegalensis* abundance influence yield across treatment

The goal here is to see the effect of locust count on end of season year. 

I will try to go across survey (locust sampling periods) and fertilizer treatment


## Library import and reading in data

```{r reading in packages, warning=FALSE,message=FALSE}


# Define a function to check and install packages if not already installed
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  }
}



# List of packages to install and load
packages <- c("tidyverse", "here", "janitor",  
              "ggpubr", "MetBrewer", "patchwork",'sf',
              'rnaturalearth','rnaturalearthdata','mgcv','gratia','emmeans')

# Call the function with your package list
install_and_load(packages)


options(width=200)
here::i_am('README.md')

```


## Correlating yield and locust count

So Im thinking this formula:

yield ~ ose_num_adj + (1|survey)

where `ose_num_adj` is locust count (fixed continuous variable) and `survey` as a random effect to help with puesdoreplication

```{r}
dat <- read_csv('data/analysis_ready_data.csv')

yield_locust_dat <- dat |> 
  select(farmer_num,village,treatment,att_mil_adj,att_mil,yield,ose_num_adj,survey) |> 
  group_by(village,farmer_num,treatment) |>
  fill(yield,.direction='up') |> 
  filter(!(survey %in% c('s1','s4'))) |> 
  drop_na(yield,ose_num_adj) 

```


lets demonstrate the correlation between yield and locust count

I've removed s1 because its before fertilization.

```{r co-correlation between treatment, yield, and locust abundance/attack}

# lcoust abundance/attack and yield correlation

raw_ose_count_yield_correlation <- yield_locust_dat |>
  ggplot(aes(x=ose_num_adj,y=yield,color=survey)) +
    geom_point() +
    theme_pubr() +
    xlab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    scale_color_met_d(name='Demuth')


# locust abundance and attack correlation

ose_count_damage_correlation <- yield_locust_dat |>
  ggplot(aes(x=ose_num_adj,y=att_mil_adj,color=treatment)) +
    geom_point() +
    theme_pubr() +
    scale_color_met_d(name='Degas')  +
    xlab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    ylab("locust damage")



# yield, abudance and attack with treatment:

p1 <- yield_locust_dat |>
  ggplot(aes(x=treatment,y=yield,color=village)) +
    geom_jitter(width=0.3,height=0) +
    theme_pubr() +
    xlab("") +
    ylab("yield") +
    scale_color_met_d(name='Demuth')

p2 <- yield_locust_dat |>
  ggplot(aes(x=treatment,y=ose_num_adj,color=village)) +
    geom_jitter(width=0.3,height=0)  +
    theme_pubr() +
    xlab("") +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    scale_color_met_d(name='Demuth')

p3 <- yield_locust_dat |>
  ggplot(aes(x=treatment,y=att_mil_adj,color=village)) +
    geom_jitter(width=0.3,height=0)  +
    theme_pubr()  +
    xlab("") +
    ylab("locust damage") +
    scale_color_met_d(name='Demuth')


(p1 + p2) / (p3 + plot_spacer())



ggsave(raw_ose_count_yield_correlation,
       file='output/locust_yield_plots/raw_ose_count_yield_correlation.png',
       width=5,height=5,dpi=300)

ggsave(ose_count_damage_correlation,
       file='output/locust_yield_plots/ose_count_damage_correlation.png',
       width=5,height=5,dpi=300)


```


### modeling yield x locust count

I am going to now model yield


```{r locust abundance/attack and yield modeling}

yield_locust_dat <- yield_locust_dat |>
  mutate(survey = factor(survey),
    treatment = factor(treatment),
    village = factor(village),
    farmer_num = factor(farmer_num))

yield_locust_mod0 <- gam(yield ~ 1,data=yield_locust_dat,family=tw())

yield_locust_mod1 <- gam(yield ~ 
    s(ose_num_adj,k=25),
  select=TRUE,
  data=yield_locust_dat,family=tw())

yield_locust_mod2 <- gam(yield ~ 
    s(ose_num_adj,k=25) + 
    s(farmer_num,bs=c('re')),
  select=TRUE,
  data=yield_locust_dat,family=tw())

yield_locust_mod3 <- gam(yield ~ 
    s(ose_num,k=25) + 
    s(farmer_num,bs=c('re')) + 
    s(treatment,bs=c('re')),
  select=TRUE,
  data=yield_locust_dat,family=tw())


# AIC model selection criteria 

AIC(yield_locust_mod0,yield_locust_mod1,yield_locust_mod2,yield_locust_mod3) |>
  as_tibble(rownames = 'model') |>
  mutate(deltaAIC = AIC - min(AIC))

# model fit diagnostics 

appraise(yield_locust_mod3)

summary(yield_locust_mod2)
summary(yield_locust_mod3)
draw(yield_locust_mod3)


```


### saving model summary to csv


```{r}

summary(yield_locust_mod2)
# Extract parametric and smooth terms
param_table <- yield_locust_mod2 |> broom::tidy(parametric = TRUE)
smooth_table <- yield_locust_mod2 |> 
    broom::tidy(parametric = FALSE)

# Standardize column names for merging
smooth_table <- smooth_table |> 
  rename(std.error = edf, estimate = ref.df) |>  # Renaming for consistency
  mutate(std.error = NA)  # Adding missing column to match parametric table

# Combine both tables
combined_table <- bind_rows(param_table, smooth_table) |> 
  mutate(type = c(rep("Parametric", nrow(param_table)), rep("Non-parametric", nrow(smooth_table))),
    term = gsub('treatment|survey','',term),
    across(where(is.numeric), ~ round(.,3)))

write_csv(combined_table,file=here('output/model_summary_files/yield_locust_abundance_model_results.csv'))


```



### creating modeled estimates, ploting, and save to disk

```{r}

mod2_ests <- gratia::smooth_estimates(yield_locust_mod2)


raw_yield_locust_count_plot <- yield_locust_dat |>
  ggplot(aes(x=ose_num_adj,y=yield)) +
    geom_point() +
    theme_pubr() +
    ylab('Yield (kg/ha)') +
    xlab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    scale_color_met_d(name='Demuth')

      


modeled_yield_locust_count_plot <- mod2_ests |>
    mutate(adjusted_estimate = .estimate) |> 
    filter(.smooth == 's(ose_num_adj)') |>
    ggplot(aes(x=ose_num_adj)) + 
      geom_ribbon(aes(ymin=adjusted_estimate - .se, ymax=adjusted_estimate + .se),alpha=.2) +
      geom_line(aes(y = adjusted_estimate),size=1) +
      geom_rug(data = yield_locust_dat,aes(x=ose_num_adj)) +
      theme_pubr() +
      ylab('Yield (partial estimates)')+
      xlab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
      ylim(-1,1)

yield_combined_plots <- raw_yield_locust_count_plot + modeled_yield_locust_count_plot + plot_annotation(tag_levels = 'a')

ggsave(yield_combined_plots,
       file='output/locust_yield_plots/ose_num.png',
       width=10,height=5,dpi=300)

ggsave(yield_combined_plots,
      file='output/locust_yield_plots/ose_num.pdf',
      width=10,height=5,dpi=300)

```

## locust count to temperature


```{r}

locust_temp_dat <- dat |> 
  select(farmer_num,village,treatment,temp_c,ose_num_adj,survey) |>
  filter(survey != 's4') |>
  drop_na(ose_num_adj,temp_c) |>
  mutate(survey = factor(survey))


```


```{r}

locust_temp_dat

temp_locust_mod0 <- gam(ose_num_adj ~ 1,
  data=locust_temp_dat,family=tw())

temp_locust_mod1 <- gam(ose_num_adj ~ 
    s(temp_c,k=25),
  select=TRUE,
  data=locust_temp_dat,family=tw())


# AIC model selection criteria 

AIC(temp_locust_mod0,temp_locust_mod1,temp_locust_mod2) |>
  as_tibble(rownames = 'model') |>
  mutate(deltaAIC = AIC - min(AIC))

summary(temp_locust_mod1)
appraise(temp_locust_mod1)

```

there is not a significant relationship between locust count and temperature

```{r}


raw_temp_locust_plot <- locust_temp_dat |>
  ggplot(aes(x=temp_c,y=ose_num_adj)) +
    geom_point()  +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^2))) +
    xlab('temperature (C)') +
    theme_pubr() 
 
temp_mod_resuls <- emmeans::emmeans(temp_locust_mod2,~temp_c)

modeled_temp_locust_plot <- temp_mod_resuls |>
  filter(.smooth == 's(temp_c)') |>
  ggplot(aes(x=temp_c)) + 
    geom_ribbon(aes(ymin=.estimate - .se, ymax = .estimate + .se),alpha=.2) +
    geom_line(aes(y = .estimate),size=1) +
    geom_rug(data = locust_temp_dat,aes(x=temp_c)) +
    theme_pubr() +
    xlab('temperature (C)') +
    ylab('locust count (partial estimates)') +
    ylim(-0.5,0.5)

temp_combined_plots <- raw_temp_locust_plot + modeled_temp_locust_plot  + plot_annotation(tag_levels = 'a')

ggsave(temp_combined_plots,
       file='output/locust_yield_plots/modeled_locust_temp_plot.png',
       width=10,height=5,dpi=300)

```

## Temperature and fertilizer plots


```{r}

temp_fert_dat <- dat |>
  select(farmer_num,village,survey,treatment,temp_c) |>
  filter(survey != 's4') |>
  drop_na()

temp_fert_plot <- temp_fert_dat |>
  ggplot(aes(x=treatment,y=temp_c,color=survey)) +
    geom_jitter(width=0.2) +
    theme_pubr() +
    scale_color_met_d(name = 'Egypt') +
    xlab('') +
    ylab('temperature (C)')


ggsave(temp_fert_plot,
  file=here('output/raw_visualization/temperature_fertilizer_plot.png'),
  width = 5,
  height = 5,
  dpi = 300)

```

## Village temperature and precipitation data


lets read in the data and manage it slightly

```{r}
weather_dat <- read_csv(here('data/processed/ERA5_Weather_Data_Per_Point.csv'),show_col_types = FALSE)

weather_dat2 <- weather_dat |>
  mutate(location = case_when(location == 'Point1' ~ 'Gniby',
    location == 'Point2' ~ 'Gossas'),
    doy = yday(date)) |>
  select(location,doy,maximum_2m_air_temperature,mean_2m_air_temperature,minimum_2m_air_temperature,total_precipitation) |>
  rename(max_temp = maximum_2m_air_temperature,
    mean_temp = mean_2m_air_temperature,
    min_temp = minimum_2m_air_temperature,
    total_precip = total_precipitation)  %>%
  group_by(location) %>%
  arrange(doy, .by_group = TRUE) %>%
  mutate(cum_precip = cumsum(total_precip)) |>
  ungroup()

```

now creating temperature and preciptation plots



```{r}


weather_dat2 |>
  select(location,doy,max_temp,mean_temp,min_temp) |>
  pivot_longer(!c(location,doy)) |>
  mutate(value = value - 273.15) |>
  ggplot(aes(x=doy,y=value,color = name,linetype=location)) +
    geom_line() +
    ggpubr::theme_pubr(legend = 'bottom') +
    MetBrewer::scale_colour_met_d(names='Peru1')

```
