
## Question 3: Does *O. senegalensis* abundance and attack influence yield across treatment

Okay so this is a tricky question to ask due to the co-correlation between treatment, yield, and locust abundance. Fertilization obviously influences both yield and locust abundance/attack. So how do we model the effect locusts have on yield *independently*?

lets demonstrate this co-correlation

```{r co-correlation between treatment, yield, and locust abundance/attack}

# lcoust abundance/attack and yield correlation

raw_ose_count_yield_correlation <- dat |>
  ggplot(aes(x=ose_num,y=yield)) +
    geom_point() +
    theme_pubr() +
    xlab("locust count")

raw_ose_damage_yield_correlation <- dat |>
  ggplot(aes(x=att_mil,y=yield)) +
    geom_point() +
    theme_pubr() +
    xlab('locust damage')


# locust abundance and attack correlation

ose_count_damage_correlation <- dat |>
  ggplot(aes(x=ose_num,y=att_mil,color=treatment)) +
    geom_point() +
    theme_pubr() +
    scale_color_met_d(name='Degas')  +
    xlab("locust count")  +
    ylab("locust damage")



# yield, abudance and attack with treatment:

p1 <- dat |>
  ggplot(aes(x=treatment,y=yield,color=village)) +
    geom_jitter(width=0.3,height=0) +
    theme_pubr() +
    xlab("") +
    ylab("yield") +
    scale_color_met_d(name='Demuth')

p2 <- dat |>
  ggplot(aes(x=treatment,y=ose_num,color=village)) +
    geom_jitter(width=0.3,height=0)  +
    theme_pubr() +
    xlab("") +
    ylab("locust count") +
    scale_color_met_d(name='Demuth')

p3 <- dat |>
  ggplot(aes(x=treatment,y=att_mil,color=village)) +
    geom_jitter(width=0.3,height=0)  +
    theme_pubr()  +
    xlab("") +
    ylab("locust damage") +
    scale_color_met_d(name='Demuth')


(p1 + p2) / (p3 + plot_spacer())



ggsave(raw_ose_count_yield_correlation,
       file='output/locust_yield_plots/raw_ose_count_yield_correlation.png',
       width=5,height=5,dpi=300)

ggsave(ose_count_damage_correlation,
       file='output/locust_yield_plots/ose_count_damage_correlation.png',
       width=5,height=5,dpi=300)


```

okay I'm not totally confident to have a model where yield, treatment, and locust abundance/attack are all included. I think theres a lot of co-correlations. So, lets just focus on locust abundance/attack and yield directly. Here is the starting of my modeling but I didnt go further....

```{r locust abundance/attack and yield modeling}

yield_locust_dat

yield_locust_mod0 <- gam(yield ~ 1,data=dat,family=tw())

yield_locust_mod1 <- gam(yield ~ ose_num,data=dat,family=tw())

yield_locust_mod2 <- gam(yield ~ ose_num +
                             s(mission,bs='re'),
                         data=dat,family=tw())



summary(yield_locust_mod1)
appraise(yield_locust_mod2)
summary(yield_locust_mod2)
```
