---
title: "Manuscript figure and table construction"
author: "Douglas Lawton"
format: html
editor: source
---

```{r reading in packages, warning=FALSE,message=FALSE}


# Define a function to check and install packages if not already installed
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  }
}



# List of packages to install and load
packages <- c("tidyverse", "here", "janitor",  
              "ggpubr", "MetBrewer", "patchwork",'sf',
              'rnaturalearth','rnaturalearthdata','mgcv','gratia','emmeans')

# Call the function with your package list
install_and_load(packages)

options(width=200)
here::i_am('README.md')
```


# Read in data


```{r}

abun_mil_attack <- read_csv(here('data/processed/locust_abundance_and_att_mil_raw_and_model_data.csv'))
yield <- read_csv(here('data/processed/yield_raw_and_model_data.csv'))

all_missiom_data <- read_csv('data/analysis_ready_data.csv') |>
  mutate(hour = lubridate::hour(datetime))

```


# yield, attack, and abundance panel
 - panel a = raw locust count through survey
 - panel b = modeled locust count through survey
 - panel c = raw locust damage through survey
 - panel d = modeled locust damage through survey
 - panel e = raw yield between treatment
 - panel f = modeled yield between treatment


 ```{r}

pA <- all_missiom_data |>
    select(village,treatment,ose_num_adj,survey) |>
    mutate(treatment = factor(case_when(survey == 's1' ~ 'PT', TRUE ~ treatment),levels=c('PT','control','fertilized')),
        survey = case_when(survey == 's1' ~ 'PT',survey == 's2' ~ 'S1',survey == 's3' ~ 'S2')) |>
    drop_na(survey) |>
    ggplot(aes(x=survey,y=ose_num_adj,color=treatment)) +
        geom_point(position = position_jitterdodge(jitter.height=0,jitter.width=0.25)) +
    scale_color_manual(values = c('dark grey','black','dark green')) +
    theme_pubr()  + 
    theme(legend.position="none")  +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^-2)))

pB <- abun_mil_attack |>
    select(treatment,starts_with('abundance_')) |>
    distinct() |>
    ggplot(aes(x=treatment,y=abundance_response,color=treatment)) +
      geom_point(size=5) +
      geom_errorbar(aes(ymin=abundance_response-abundance_se,ymax=abundance_response+abundance_se),width=0) +
      scale_color_manual(values = c('black','dark green')) +
      theme_pubr()  + 
      theme(legend.position="none") +
      xlab('') +
      ylab('') +
  theme(axis.text.x=element_blank())


pC <- all_missiom_data |>
    select(village,treatment,att_mil_adj,survey) |>
    mutate(treatment = factor(case_when(survey == 's1' ~ 'PT', TRUE ~ treatment),levels=c('PT','control','fertilized')),
        survey = case_when(survey == 's1' ~ 'PT',survey == 's2' ~ 'S1',survey == 's3' ~ 'S3')) |>
    drop_na(survey) |>
    ggplot(aes(x=survey,y=att_mil_adj,color=treatment)) +
        geom_point(position = position_jitterdodge(jitter.height=0,jitter.width=0.25)) +
    scale_color_manual(values = c('dark grey','black','dark green')) +
    theme_pubr()  + 
    theme(legend.position="none") +
    xlab('') +
    ylab('locust Damage')

pD <- abun_mil_attack |>
    select(treatment,starts_with('att_mil_')) |>
    distinct() |>
    ggplot(aes(x=treatment,y=att_mil_response,color=treatment)) +
      geom_point(size=5) +
      geom_errorbar(aes(ymin=att_mil_response-att_mil_se,ymax=att_mil_response+att_mil_se),width=0) +
      scale_color_manual(values = c('black','dark green')) +
      theme_pubr()  + 
      theme(legend.position="none") +
      xlab('') +
      ylab('') +
  theme(axis.text.x=element_blank())

pE <- all_missiom_data |>
    select(village,treatment,yield) |>
    distinct() |>
    mutate(treatment = factor(treatment,levels=c('PT','control','fertilized'))) |>
    ggplot(aes(x=treatment,y=yield,color=treatment)) +
        geom_point(position = position_jitterdodge(jitter.height=0,jitter.width=0.25)) +
    scale_color_manual(values = c('black','dark green')) +
    theme_pubr()  + 
    theme(legend.position="none") +
    xlab('') +
    ylab('yield')

pF <- yield |>
    filter(model == 'GLMM') |>
    select(treatment,response,SE) |>
    distinct() |>
    group_by(treatment) |>
    mutate(reponse = mean(response),SE = mean(SE)) |>
    ggplot(aes(x=treatment,y=reponse,color=treatment)) +
      geom_point(size=5) +
      geom_errorbar(aes(ymin=reponse-SE,ymax=reponse+SE),width=0) +
      scale_color_manual(values = c('black','dark green')) +
      theme_pubr()  + 
      theme(legend.position="none") +
      xlab('') +
      ylab('')




 ```


 ```{r}
 
 design <- ("
  1#2
  3#4
  5#6
")

manuscript_figure <- pA + pB + pC + pD + pE + pF + plot_layout(design = design)

ggsave(manuscript_figure,file=here('output/panel_1.pdf'),width=8,height=8)


 ```



## option 2: both model and raw data at the same scale

# yield, attack, and abundance panel
 - panel a = raw locust count through survey
 - panel b = modeled locust count through survey
 - panel c = raw locust damage through survey
 - panel d = modeled locust damage through survey
 - panel e = raw yield between treatment
 - panel f = modeled yield between treatment


 ```{r}

pA_opt2 <- all_missiom_data |>
    select(village,treatment,ose_num_adj,survey) |>
    mutate(treatment = factor(case_when(survey == 's1' ~ 'PT', TRUE ~ treatment),levels=c('PT','control','fertilized')),
        survey = case_when(survey == 's1' ~ 'PT',survey == 's2' ~ 'S1',survey == 's3' ~ 'S2')) |>
    drop_na(survey) |>
    ggplot(aes(x=survey,y=ose_num_adj,color=treatment)) +
        geom_point(position = position_jitterdodge(jitter.height=0,jitter.width=0.25)) +
    scale_color_manual(values = c('dark grey','black','dark green')) +
    theme_pubr()  + 
    theme(legend.position="none")  +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^-2))) +
    ylim(0,90)

pB_opt2 <- abun_mil_attack |>
    select(treatment,starts_with('abundance_')) |>
    distinct() |>
    ggplot(aes(x=treatment,y=abundance_response,color=treatment)) +
      geom_point(size=2) +
      geom_errorbar(aes(ymin=abundance_response-abundance_se,ymax=abundance_response+abundance_se),width=0) +
      scale_color_manual(values = c('black','dark green')) +
      theme_pubr()  + 
      theme(legend.position="none") +
      xlab('') +
      ylab('') +
      theme(axis.text.x=element_blank())  +
      ylim(0,90)


pC_opt2 <- all_missiom_data |>
    select(village,treatment,att_mil_adj,survey) |>
    mutate(treatment = factor(case_when(survey == 's1' ~ 'PT', TRUE ~ treatment),levels=c('PT','control','fertilized')),
        survey = case_when(survey == 's1' ~ 'PT',survey == 's2' ~ 'S1',survey == 's3' ~ 'S3')) |>
    drop_na(survey) |>
    ggplot(aes(x=survey,y=att_mil_adj,color=treatment)) +
        geom_point(position = position_jitterdodge(jitter.height=0,jitter.width=0.25)) +
    scale_color_manual(values = c('dark grey','black','dark green')) +
    theme_pubr()  + 
    theme(legend.position="none") +
    xlab('') +
    ylab('locust Damage') + 
    ylim(0,0.5)

pD_opt2 <- abun_mil_attack |>
    select(treatment,starts_with('att_mil_')) |>
    distinct() |>
    ggplot(aes(x=treatment,y=att_mil_response,color=treatment)) +
      geom_point(size=2) +
      geom_errorbar(aes(ymin=att_mil_response-att_mil_se,ymax=att_mil_response+att_mil_se),width=0) +
      scale_color_manual(values = c('black','dark green')) +
      theme_pubr()  + 
      theme(legend.position="none") +
      xlab('') +
      ylab('') +
      theme(axis.text.x=element_blank()) + 
      ylim(0,0.5)

pE_opt2 <- all_missiom_data |>
    select(village,treatment,yield) |>
    distinct() |>
    mutate(treatment = factor(treatment,levels=c('PT','control','fertilized'))) |>
    ggplot(aes(x=treatment,y=yield,color=treatment)) +
        geom_point(position = position_jitterdodge(jitter.height=0,jitter.width=0.25)) +
    scale_color_manual(values = c('black','dark green')) +
    theme_pubr()  + 
    theme(legend.position="none") +
    xlab('') +
    ylab('yield')  + 
    ylim(0,2100)

pF_opt2 <- yield |>
    filter(model == 'GLMM') |>
    select(treatment,response,SE) |>
    distinct() |>
    group_by(treatment) |>
    mutate(reponse = mean(response),SE = mean(SE)) |>
    ggplot(aes(x=treatment,y=reponse,color=treatment)) +
      geom_point(size=2) +
      geom_errorbar(aes(ymin=reponse-SE,ymax=reponse+SE),width=0) +
      scale_color_manual(values = c('black','dark green')) +
      theme_pubr()  + 
      theme(legend.position="none") +
      xlab('') +
      ylab('')  + 
      ylim(0,2100)




 ```


 ```{r}
 
 design <- ("
  1#2
  3#4
  5#6
")

manuscript_figure_opt2 <- pA_opt2 + pB_opt2 + pC_opt2 + pD_opt2 + pE_opt2 + pF_opt2 + plot_layout(design = design)

ggsave(manuscript_figure_opt2,file=here('output/panel_1_opt2.png'),width=8,height=8)


 ```


# supplementary figures


## saving out as png

I am going to save those plots out int `output/raw_visualization`


```{r}

# Define directory
output_dir <- here("output", "raw_visualization")

# Create plots with better axis labels
plots <- list(
  "scatter_treatment_ose_num" = dat |>
    filter(survey != 's1') |>
    ggplot(aes(x = treatment, y = ose_num_adj)) +
    geom_jitter(width = 0.2, height = 0) +
    labs(x = "treatment") +
    ylab(bquote("Locust count" ~ (individuals %*% 100 ~ m^-2))) +
    theme_pubr(),
  
  "scatter_treatment_att_mil" = dat |>
    filter(survey != 's1') |>
    ggplot(aes(x = treatment, y = att_mil_adj)) +
    geom_jitter(width = 0.2, height = 0) +
    labs(x = "treatment", y = "locust damage") +
    theme_pubr(),

  "scatter_treatment_yield" = dat |>
    filter(survey != 's1') |>
    ggplot(aes(x = treatment, y = yield)) +
    geom_jitter(width = 0.2, height = 0) +
    labs(x = "treatment", y = "yield") +
    theme_pubr(),

  "scatter_ose_num_att_mil" = dat |>
    filter(survey != 's1') |>
    ggplot(aes(x = ose_num, y = att_mil_adj)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(y = "locust damage") +
    ylab(bquote("Locust count" ~ (individuals %*% 100 ~ m^-2))) +
    theme_pubr(),

  "scatter_ose_num_temp" = dat |>
    ggplot(aes(y = ose_num_adj, x = temp_c)) +
    geom_point() +
    labs(x = "temperature (°C)") +
    ylab(bquote("Locust count" ~ (individuals %*% 100 ~ m^-2))) +
    theme_pubr(),

  "scatter_ose_num_humidity" = dat |>
    ggplot(aes(y = ose_num_adj, x = humidity_percent)) +
    geom_point() +
    labs(x = "humidity (%)") +
    ylab(bquote("Locust count" ~ (individuals %*% 100 ~ m^-2))) +
    theme_pubr(),

  "scatter_ose_num_ground_cover" = dat |>
    ggplot(aes(y = ose_num_adj, x = ground_cover_percent)) +
    geom_point() +
    labs(x = "ground cover (%)") +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^-2))) +
    theme_pubr(),

  "scatter_ose_num_survey_village" = dat  |>
    filter(survey != 's4') |>
    ggplot(aes(y = ose_num_adj, x = survey, color = village)) +
    geom_jitter(width = 0.3, height = 0) +
    MetBrewer::scale_color_met_d(name = "Degas") +
    labs(x = "survey", color = "village") +
    ylab(bquote("locust count" ~ (individuals %*% 100 ~ m^-2))) +
    theme_pubr(),

  "scatter_att_mil_survey_village" = dat |>
    filter(survey != 's4') |>
    ggplot(aes(y = att_mil_adj, x = survey, color = village)) +
    geom_jitter(width = 0.3, height = 0) +
    MetBrewer::scale_color_met_d(name = "Degas") +
    labs(x = "survey", y = "locust damage", color = "village") +
    theme_pubr()
)


walk2(names(plots), plots, ~ ggsave(
  filename = file.path(output_dir, paste0(.x, ".png")), # Save as PNG
  plot = .y,
  width = 5, height = 5, dpi = 300
))

# Spatial plots
africa <- ne_countries(scale = "medium", continent = "Africa", returnclass = "sf") |>
  select(name_en,geometry)

senegal <- africa |> filter(name_en == "Senegal")

plot(senegal)

# Create city points
city_data <- data.frame(
  name = c("Gniby", "Gossas"),
  latitude = c(14.460248196957838, 14.533427928227212),
  longitude = c(-15.682083368564234, -16.079953061378422)
)

# Convert city data to sf object
cities <- st_as_sf(city_data, coords = c("longitude", "latitude"), crs = 4326)

# Plot
senegal_points <- ggplot() +
  geom_sf(data = senegal, color = "black",fill='light grey') +  # Senegal boundary
  #geom_sf(data = hulls, alpha = 0.3, aes(fill=village,color=village)) +  # Convex hulls
  geom_sf(data = cities, color = "black", size = 3, shape = 21,aes(fill=name)) +  # City markers
  ggrepel::geom_text_repel(data = cities, aes(label = name, geometry = geometry), stat = "sf_coordinates", nudge_y = 0.5, size = 4) +  # City labels
  theme_void() +
  MetBrewer::scale_color_met_d(name='Archambault')  +
  MetBrewer::scale_fill_met_d(name='Archambault') + 
  theme(legend.position="none")

africa_map <- ggplot()  +
  geom_sf(data=africa,fill='light grey',color= "grey") +
  geom_sf(data = senegal, color = "black",fill='grey') +
  theme_void()


 design <- ("
  112
  112
")

map <- senegal_points + africa_map + plot_layout(design = design)
map


ggsave(
  filename = file.path(output_dir, paste0('city_map', ".png")), # Save as PNG
  plot = map,
  width = 5, height = 5, dpi = 300
)

```
